Decisions Decisions Decisions: Carrying out Latent Profile Analysis in Accordance With Best Practices Using the tidyLPA R package

### 2020-10-12 paper.rmd
update aper

---
title             : "Decisions Decisions Decisions: Carrying out Latent Profile Analysis in Accordance With Best Practices Using the tidyLPA R package"
shorttitle        : "tidyLPA"

@@

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
library(tidyverse)
knitr::clean_cache()
knitr::opts_chunk$set(cache.extra = knitr::rand_seed, warning = FALSE)
```

@@

```{r, include = FALSE, eval = FALSE}
library(simstudy)

def <- defData(varname = "x1", formula = 65, variance = 30)
def <- defData(def, varname = "x2", formula = 85, variance = 23)
def <- defData(def, varname = "xMix", 
               formula = "x1 | .5 + x2 | .5", 
               dist = "mixture")

@@

```{r, echo = FALSE, eval = FALSE}

ggplot(dx, aes(x = xMix)) +
  geom_density() +
  papaja::theme_apa() +
  ylim(0, .10)
```

@@

```{r, echo = FALSE, eval = FALSE}
dx %>% 
  gather(key, val, x1:x2) %>% 
  ggplot(aes(x = val, fill = key)) +
  geom_density(alpha = .3) +
  papaja::theme_apa() +
  ylim(0, .10)
```

@@

In this tutorial, we aim to describe one of the most commonly-used--and relevant
to psychologists--application of the general mixture model, to cases for which
all of the variables for which (relatively) homogeneous groups are identified
from among a (relatively) heterogeneous sample are continuos, *latent profile
analysis* (LPA). We note that scholars using the approach, especially in
developmental and educational psychology, use cluster analytic methods. Such an
approach has some similarities with LPA (CITE paper showing how k-means is
similar to the simplest model specification of LPA?), though LPA also has a
number of benefits (Linnenbrink-Garcia & Wormington, 2017), and we we focus on
LPA in this paper.

@@

The goal of LPA is estimate the parameters for a number of distributions
(typically multivariate) from a single dataset. Thus, such an approach is
model-based, and some descriptions in the literature refer to it as model-based
clustering (Hennig, Meila, Murtagh, & Rocci, 2015; Scrucca, Fop, Murphy, &
Raftery, 2017). Thus, one distinction between LPA and other, similar cluster
analytic approaches is that LPA is model-based; instead of using algorithms to
group together cases, LPA seeks to estimate parameters - in terms of variances
and covariances and how they are the same or different across profiles - that
best characterize the different distributions. Then, this approach seeks to
assign to each observation a probability that the observation is a sample from
the population associated with each profile (or mixture component).

@@

We wish to note that challenge facing the analyst using LPA is that these
parameters and the distinct model parameterizations that can be estimated is the
different terminology used. As one example, Scrucca et al. (2017) refer to these
parameterizations not in terms of whether and how parameters are estimated, but
rather in terms of the geometric properties of the distributions that result
from particular parameterizations. Muthen and Muthen (1997-2017) and others
(Pastor et al., 2007) commonly refer to local independence to mean that the
covariances are fixed to zero (also described as the specification of the
covariance matrix as “diagonal,” because only the diagonal components, or the
variances, are estimated).

@@

## Other Tools

@@

You can install tidyLPA from CRAN with:

@@

You can also install the development version of tidyLPA from GitHub with:

@@

``` {r, message = F}
library(tidyLPA)
library(dplyr)
```

@@

Here, we use a simulated dataset, based on Curry et al., 2019, contains data on moral relevance and judgment across the seven domains of the *Morality As Cooperation* scale.

@@

We can see a number of statistics for each of the models that were estimated, including the 
AIC (Aikake information criterion; based on -2 log-likelihood, and penalized by number of parameters) and BIC (BIC: Bayesian information criterion; based on -2 log-likelihood, and penalized by number of parameters adjusted by sample size). Both are _penalized_ likelihood statistics; the log-likelihood will always decrease with the addition of additional 
paremeters estimated by the model, whereas the AIC and BIC will account for the number of parameters; lower values of
either indicate better fit.

@@

```{r, get-fit-m1}
get_fit(m1)
```

@@

Here is an example of specifying a model with varying variances and covariances
(Model 6; not run here):

``` {r, curry-mac-m6, eval = F}
curry_mac %>% 
  select(KinshipR:PossessionR, 
         KinshipJ:PossessionJ) %>% 
  estimate_profiles(3, 
                    variances = "varying",
                    covariances = "varying")
```

@@

````{r, compare-mclust, eval= FALSE}
compare_solutions_mclust <- curry_mac %>% 
  select(KinshipJ:PossessionJ) %>% 
  estimate_profiles(models = c(1, 2, 3, 6),
                    n_profiles = 1:9,
                    package = "mclust")
```

@@

```{r, compare-mplus, eval = FALSE}
compare_solutions_mplus <- curry_mac %>% 
  select(KinshipJ:PossessionJ) %>% 
  estimate_profiles(models = c(1, 2, 3, 4, 5, 6),
                    n_profiles = 1:9,
                    package = "mplus")
```

@@

```{r, compare-mplus-starts, eval = FALSE}
compare_solutions_mplus_starts <- curry_mac %>% 
  select(KinshipJ:PossessionJ) %>% 
  estimate_profiles(models = c(1, 2, 3, 4, 5, 6),
                    n_profiles = 1:9,
                    package = "mplus",
                    ANALYSIS = "starts = 100, 20;")
```

@@

@@

```{r, echo = FALSE}
t <- read_csv("GRoLTS-applicability-to-LPA.csv")

>>>>

---
title             : "Decisions Decisions Decisions: Carrying out Latent Profile Analysis in Accordance With Best Practices Using the tidyLPA R package"
shorttitle        : "LPA Best Practices"

@@

```{r analysis-preferences, echo = FALSE}
# Seed for random number generation
set.seed(42)
library(tidyverse)
library(simstudy)

knitr::clean_cache()
knitr::opts_chunk$set(cache.extra = knitr::rand_seed, warning = FALSE)
```

@@

```{r, include = FALSE, eval = TRUE}
def <- defData(varname = "x1", formula = 65, variance = 30)
def <- defData(def, varname = "x2", formula = 85, variance = 23)
def <- defData(def, varname = "xMix", 
               formula = "x1 | .5 + x2 | .5", 
               dist = "mixture")

@@

```{r, include = FALSE, eval = TRUE}
ggplot(dx, aes(x = xMix)) +
  geom_density() +
  papaja::theme_apa() +
  ylim(0, .10)
```

@@

```{r, include = FALSE, eval = TRUE}
dx %>% 
  gather(key, val, x1:x2) %>% 
  ggplot(aes(x = val, fill = key)) +
  geom_density(alpha = .3) +
  papaja::theme_apa() +
  ylim(0, .10)
```

@@

In this tutorial, we aim to describe one of the most commonly-used--and relevant
to psychologists---application of the general mixture model, to cases for which
all of the variables for which (relatively) homogeneous groups are identified
from among a (relatively) heterogeneous sample are continuos, *latent profile
analysis* (LPA). We note that scholars using the approach, especially in
developmental and educational psychology, use cluster analytic methods. Such an
approach has some similarities with LPA (CITE paper showing how k-means is
similar to the simplest model specification of LPA?), though LPA also has a
number of benefits (Linnenbrink-Garcia & Wormington, 2017), and we we focus on
LPA in this paper.

@@

The goal of LPA is estimate the parameters for a number of distributions
(typically multivariate) from a single data set. Thus, such an approach is
model-based, and some descriptions in the literature refer to it as model-based
clustering (Hennig, Meila, Murtagh, & Rocci, 2015; Scrucca, Fop, Murphy, &
Raftery, 2017). Thus, one distinction between LPA and other, similar cluster
analytic approaches is that LPA is model-based; instead of using algorithms to
group together cases, LPA seeks to estimate parameters - in terms of variances
and covariances and how they are the same or different across profiles - that
best characterize the different distributions. Then, this approach seeks to
assign to each observation a probability that the observation is a sample from
the population associated with each profile (or mixture component).

@@

We note that challenge facing the analyst using LPA is that these
parameters and the distinct model parameterizations that can be estimated is the
different terminology used. As one example, Scrucca et al. (2017) refer to these
parameterizations not in terms of whether and how parameters are estimated, but
rather in terms of the geometric properties of the distributions that result
from particular parameterizations. Muthen and Muthen (1997-2017) and others
(Pastor et al., 2007) commonly refer to local independence to mean that the
covariances are fixed to zero (also described as the specification of the
covariance matrix as “diagonal,” because only the diagonal components, or the
variances, are estimated).

@@

## Tools other than tidyLPA

@@

You can install tidyLPA from the Comprehensive R Archive Network, or CRAN, with:

@@

This is how we recommend most users install tidyLPA.

You can also install the development version of tidyLPA, which may have new features that are not yet availalble on CRAN (because the version on CRAN is periodically updated) from GitHub with:

@@

``` {r, message = F}
library(tidyLPA)
library(dplyr) # for preparing the data
```

@@

Here, we use a simulated dataset contains data on moral relevance and judgment across the seven domains of the *Morality As Cooperation* scale. More on this scale can be found in Curry et al., (2019).

Add one or two sentences here on the dataset?

@@

We can see a number of statistics for each of the models that were estimated, including the 
AIC (Aikake information criterion; based on -2 * the log-likelihood, and penalized by number of parameters) and BIC (BIC: Bayesian information criterion; based on -2 log-likelihood, and penalized by number of parameters adjusted by sample size). Both are _penalized_ likelihood statistics; the log-likelihood will always decrease with the addition of additional 
paremeters estimated by the model, whereas the AIC and BIC will account for the number of parameters; lower values of
either indicate better fit.

@@

```{r, get-fit-m1}
get_fit(m1) %>% 
  knitr::kable()
```

@@

Here is an example of specifying a model with varying variances and covariances
(Model 6)

``` {r, curry-mac-m6, eval = TRUE}
curry_mac %>% 
  select(KinshipR:PossessionR, 
         KinshipJ:PossessionJ) %>% 
  estimate_profiles(3, 
                    variances = "varying",
                    covariances = "varying")
```

@@

````{r, compare-mclust, eval= TRUE}
compare_solutions_mclust <- curry_mac %>% 
  select(KinshipJ:PossessionJ) %>% 
  estimate_profiles(models = c(1, 2, 3, 6),
                    n_profiles = 1:9,
                    package = "mclust") %>% 
  knitr::kable()
```

@@

```{r, compare-mplus, eval = FALSE}
compare_solutions_mplus <- curry_mac %>% 
  select(KinshipJ:PossessionJ) %>% 
  estimate_profiles(models = c(1, 2, 3, 4, 5, 6),
                    n_profiles = 1:9,
                    package = "mplus") %>% 
  knitr::kable()
```

@@

```{r, compare-mplus-starts, eval = TRUE}
compare_solutions_mplus_starts <- curry_mac %>% 
  select(KinshipJ:PossessionJ) %>% 
  estimate_profiles(models = c(1, 2, 3, 4, 5, 6),
                    n_profiles = 1:9,
                    package = "mplus",
                    ANALYSIS = "starts = 100, 20;") %>% 
  knitr::kable()
```

@@

The GRoLTs checklist is a helpful way to think about how best practices for LPA;
here, we show how to implement these using tidyLPA. First, here is the checklist:

@@

```{r, echo = FALSE}
t <- readr::read_csv("GRoLTS-applicability-to-LPA.csv")
